Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
fix breakage

\subsection{Proofs of Templates} \
%than{Can someone suggest another title?}
In this section, we verify \lstinline{traverse} for both lock-coupling and give-up templates.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ (\texttt{k} \in \texttt{range})\ \land \ \nodeboxrep(\texttt{pn->n'})  \ \ast \  \texttt{R(pn->n')}   
\end{mathpar}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{m}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\texttt{(res} \leftrightarrow (\texttt{pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

\begin{mathpar}
	\centering
{\color{blue} \langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }

\texttt{insert(p, k, v)}

{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}\rangle}

{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \rangle }

 \texttt{lookup(p, k)}  \

{\color{blue}\langle \texttt{v}.\ \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{m} \land \texttt{m(k) = v}\rangle }

{\color{blue}	\langle \texttt{m}.\ \nodeboxrep\ \texttt{p} \ |\ \treerep\ \texttt{m}\rangle }

\texttt{delete(p, k)} \ \ \ \ \ \ \

{\color{blue} \langle \nodeboxrep\ \texttt{p}\ |\ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}\rangle }
\end{mathpar}

@@

The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}). In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp(\texttt{pn->n})$ and the public precondition $\treerep\ \texttt{m}$. Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.

Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}. Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ (\texttt{m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.

>>>>

\subsection{Proofs of Templates} 
In this section, we present our verification methodology, termed \lstinline{traverse}, which applies to both lock-coupling and give-up templates. The detailed exploration of these template verifications will be discussed in Section \ref{traverse_proof_lock} and \ref{traverse_proof_giveup}. Before we delve into the intricacies of these verifications, it is essential to familiarize ourselves with the concept of \emph{atomic triples}.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			 \forall \  \texttt{m}. \left\langle
				\nodeboxrep \texttt{(pn->n)}  \ \ast \ \texttt{R(pn->n)} \ \big| \ \treerep\ \texttt{m} \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\ \texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \ \nodeboxrep \texttt{(pn->n')} \\  
				\ast \ \texttt{R(pn->n')} \ast \ \texttt{(k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template}
	\label{fig:traverse_lock}
\end{figure}

@@

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
\end{mathpar}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  \texttt{m}. \left\langle
			\begin{array}{c}
				\infp (\texttt{pn->n}) \ \big| \ \treerep\ \texttt{m}
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \exists \  \texttt{res} \ \texttt{n'}.
			\begin{array}{c}
				\texttt{(res} \leftrightarrow \texttt{(pn->n'->t} = \texttt{NULL))}  \ \ast \infp \texttt{(pn->n')} 
				\\ 
				\ast \ \texttt{R(pn->n')} \ \ast \ (\texttt{k} \in \texttt{range(n'))}
			\end{array}
			\ \Bigg| \ \treerep\ \texttt{m} \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

@@

\begin{mathpar}
	{\color{blue}
		\texttt{m.}\left\langle 
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
		\right\rangle
	}
	\ \texttt{insert(p, k, v)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ (\texttt{m[k} \mapsto \texttt{v])}
		\right\rangle
	}
\\
	{\color{blue}
		\texttt{m.}\left\langle 
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
		\right\rangle
	}
	\ \texttt{lookup(p, k)}\ 
	{\color{blue}
		\left\langle \texttt{v}. \ \
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m} \land \texttt{m(k) = v}
		\right\rangle
	}
\\
	{\color{blue}
		\texttt{m.}\left\langle 
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{m}
		\right\rangle
	}
	\ \texttt{delete(p, k)}\ 
	{\color{blue}
		\left\langle 
		\nodeboxrep\ \texttt{p} \ \big | \ \treerep\ \texttt{(m[k} \mapsto \_ \texttt{])}
		\right\rangle
	}
\end{mathpar}

@@

The proof for the \texttt{insert} operation begins with the preconditions $\nodeboxrep \ \texttt{p}$ and $\treerep \ \texttt{m}$. We assign \texttt{pn->n} to point to the same memory address as \texttt{*p} (line 3 in Figure \ref{insert_giveup}). In Section \ref{traverse_proof_giveup}, we discuss providing the identical preconditions for the \texttt{traverse} function, which encompasses the local precondition $\infp\texttt{(pn->n)}$ and the public precondition $\treerep\ \texttt{m}$. Subsequently, we apply the \texttt{traverse} specification from Section \ref{traverse_proof_giveup}.

Analogous to the lock-coupling template's \texttt{insert} operation, there are two possible branches representing the outcome following the execution of \texttt{traverse}. Once we demonstrate the ability to atomically update the tree's abstract state from $\treerep \ \texttt{m}$ to $\treerep\ \texttt{(m[k} \mapsto \texttt{v])}$, we proceed by releasing the node \texttt{n} and returning it to the shared state, thereby finalizing the proof.
